Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory

### 2018-03-06 Cars/paper.tex
paper

Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020. These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery, and selects a path to avoid any collisions. As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self-driving vehicles. However self-driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
\\

While characteristics of surrounding vehicles, such as theor postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g. \citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory has seen less attention in the literature. These forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
\\

This paper takes a statistical approach to this trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset. It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that can lead to a collision to be detected. The time series models used allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them. Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
\\

@@

We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in the model substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous model. These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction on a prior distribution for parameters that are approximately inferred in an online setting.
\\

This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used and steps required to extract acceleration and steering angle for each vehicle, while Section \ref{sec:models} introduces several time series models used. Section \ref{sec:Inference} reviews Bayesian inference, which is extended to the online updating setting in Section \ref{sec:UVB}. Finally Section \ref{sec:eval} contains empirical results, and Section \ref{sec:disc} concludes the vehicle trajectory forecasting study and discusses future work.

\section{Data processing}
\label{sec:dataProcessing}
Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005. Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section. 
Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1. Vehicles that entered or exited midway through the freeway section, are excluded in this paper. There is a curvature to the road occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.
\\

@@

Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature. To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self-driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates. This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane. \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway. Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
\\

Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
\begin{align}
x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
\end{align}
Note that when $\delta_{i, t} \pm \pi/2$ then car $i$ takes a position parallel to the centre line, and hence $x^*_{i, t} = x^*_{i, t-1}$.
\\

From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs from driver $i$ are calculated via
\begin{align}
\delta_{i, t} &= 
     \begin{cases}
       \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right)  &\quad\text{if }x^*_{i, t} \neq x^*_{i, t-1} \\
       \frac{\pi}{2} &\quad\text{if } y^*_{i, t} > y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       -\frac{\pi}{2} &\quad\text{if } y^*_{i, t} < y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       \delta_{i, t-1} &\quad\text{otherwise} \\ 
     \end{cases} \label{dEq} \\
v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
\end{align}
In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=2,3,...,T\}$, according to
\begin{equation}
\label{aEq}
a_{i, t} = v_{i, t} - v_{i, t-1}. 
\end{equation}
\\

Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$. 3387 vehicles remain in the dataset after the above exclusion criteria are applied. The data into a two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
\\

The training set and the test set are treated as observations from vehicles at two distinct time periods. The training set represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles. The test set represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.

@@

@@

Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles. These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent heterogenous (IH) model and the hierarchical, clustered heterogeneity (CH) model. The ID model does not incorporate information from the training set of vehicles into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set. 
\\

@@

The framework provided by (\ref{xEq}) and (\ref{yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \dots, T+H\}$.  
\\

@@

These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and angle  $\delta_t$ are employed, given by
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
\end{align}
for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution. 
\\

@@

The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them. To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous (IH) model} and a dependent \textit{clustered heterogeneous} (CH) model. In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})$ corresponding to the parameters given in the model for car $i$ given in
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{aAR2} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{dAR2}
\end{align}
with $\epsilon_{i,t}$ and $\eta_{i,t}$ all independent standard normal random variables. For the IH model, each $\theta_i,$ for $i=1,2,\ldots, N$, is assumed to follow an independent $N(\mu, \Sigma)$ prior distribution, resulting in a joint posterior distribution that may be decomposed into $N$ distinct independent marginal distributions, i.e.
\begin{equation}
p(\theta_{1:N} \mid z_{1:N}) = \prod_{i=1}^N p(\theta_{i} \mid z_{i}),
\end{equation}
due to the distinct parameters and independent (albeit identical) prior distributions for each of the $N$ models. \footnote{Recall that $z_{i}=\{x^*_{i,s},y^*_{i,s},\mbox{ for } s=1,2,...,T \}$.}
\\

In contrast, the alternative CH model does not assume \textit{a priori} independence of each of the $\theta_i$ vectors and instead clusters individual $\theta_i$ values into one of $K$ common latent multivariate normal distributions as determined by $\beta.$ The hierarchical CH model assumes that $\beta$ is itself random, with $\beta \sim p(\beta)$. By augmenting each driver's individual parameter $\theta_i$ with the corresponding $k_i=j$ variable, where $j$ can take on any value $1,2 \ldots, K$, we have
\begin{equation}
\label{mixPrior}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
\end{equation}
and
\begin{equation}
k_i \mid \beta \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
\end{equation}
for each $i = 1, \dots, N$. The mixture indicator variables satisfying $k_i=j$ determine a cluster of similar $\theta_i$ parameters corresponding to the behaviours of a subset of drivers. $\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, for j=1,\},$ with \textit{a priori} assumptions for each cluster given by  
\begin{align}
\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right).
\end{align}
\\

@@

For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles that are available. We conceive of such information begin available before any self-driving vehicle enters a lone on the road. Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional behicle that may be encountered while the self-driving vehicle is on the road. Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available.
\\

@@

Combining a fast inference procedure with the test vehicle posterior distributions implied by the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.

@@

Given a series of data for vehicle $i$ observed up to time $T$, $z_{i, 1:T}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
\begin{equation}
\label{predictive}
p(z_{i, T+h} | z_{i, 1:T}) =\int p(z_{i, T+h}|z_{i, 1:T}, \theta_i) p(\theta_i | \textbf{z}_{1:N}, z_{i, 1:T}) d\theta.
\end{equation}
To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewcar2}) for the CH model must first be inferred. However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive} is unavailable.
\\

Collectively denoting all available data, for either $\textbf{z}_{1:N}$ or $z_{i}$, as $\textbf{z}$ for the remainder of this section, two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB). In brief, MCMC is used to create a sample from $p(\theta_i | \textbf{z})$, with any function of $\theta_i$ that is desired estimated from that sample. In contrast, VB replaces $p(\theta_i | \textbf{z})$ with a parametric approximation, denoted by $q_{\lambda}(\theta_i |\textbf{z})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $\textbf{z}$.

@@

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler. The Gibbs sampler algorithm iteratively samples the components of a $k$dimensional parameter vector $\theta_i$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_{i, 1} | \theta_{i, 2}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
&p(\theta_{i, 2} | \theta_{i, 1}, \theta_{i, 3}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_{i, k} | \theta_{i, 1}, \dots, \theta_{o, k-1}, \textbf{z}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta_i | \textbf{z})$. Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm. The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model. The full conditional distributions cannot be recognised for each of the time series models proposed in Section \ref{sec:models} and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
\\

@@

Eucliedean Error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}. Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error. Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3). At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model.
\\

>>>>

Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles to be available to consumers as early as 2020. These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery, and selects a path to avoid collisions. With as many as $94\%$ of accidents in the United States resulting from human error, according to the US National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, the introduction of fully self-driving vehicles is expected to substantially improve safety on the roads. However self-driving vehicles will operate in situations where the majority of surrounding vehicles are still controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
\\

While characteristics of surrounding vehicles, such as theirr postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g. \citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory has seen less attention in the literature. These forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
\\

This paper takes a statistical approach to this trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset. It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that could cause a collision to be detected. The time series models used allow for heterogeneity between different vehicles, which may result from sources such as differences in individual driving styles, or the conditions of traffic. Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration, while forecasts for more consistent drivers will benefit from a low variance.
\\

@@

We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in these time series models substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous approach. These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction of a prior and approximating distribution for parameters that are inferred in an online setting with Variational Bayes.
\\

This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used and steps required to extract acceleration and steering angle for each vehicle, while Section \ref{sec:models} introduces several time series models used. Section \ref{sec:Inference} reviews Bayesian inference, which is extended to the online, approximate updating setting in Section \ref{sec:UVB}. Finally Section \ref{sec:eval} contains empirical results, and Section \ref{sec:disc} concludes the vehicle trajectory forecasting study and discusses future work.

\section{Data processing}
\label{sec:dataProcessing}
Data is provided by the Next Generation Simulation (NGSIM) project conducted by the US Federal Highway Administration (FHWA), which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005. Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section. 
Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1. Vehicles that entered or exited midway through the freeway section are excluded in this paper. There is a curvature to the road occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.
\\

@@

Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature. To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self-driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates. This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane. \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway. Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the cente of the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
\\

Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ follows friom the trigonometric relationship in Figure \ref{fig:motion}:
\begin{align}
x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
\end{align}
Note that when $\delta_{i, t} \pm \pi/2$ then car $i$ takes a position parallel to the centre line, and hence $x^*_{i, t} = x^*_{i, t-1}$.
\\

From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs to motion from driver $i$ are calculated via
\begin{align}
\delta_{i, t} &= 
     \begin{cases}
       \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right)  &\quad\text{if }x^*_{i, t} \neq x^*_{i, t-1} \\
       \frac{\pi}{2} &\quad\text{if } y^*_{i, t} > y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       -\frac{\pi}{2} &\quad\text{if } y^*_{i, t} < y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       \delta_{i, t-1} &\quad\text{otherwise,} \\ 
     \end{cases} \label{dEq} \\
v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
\end{align}
In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=3,4,...,T\}$, according to
\begin{equation}
\label{aEq}
a_{i, t} = v_{i, t} - v_{i, t-1}. 
\end{equation}
\\

Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$. Some additional vehicles are excluded due to an insufficient number of observations, leaving 3387 vehicles in the dataset in a relative coordinate system. These vehicles are split into two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
\\

The training set and the test set are treated as observations from vehicles at two distinct time periods, illustrated in Figure \ref{fig:problem}. The training set in the left panel represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles. The test set in the right panel represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.

@@

\begin{figure}
\centering
\includegraphics[width = 0.95\textwidth]{problem}
\caption{Left: The paths of vehicles in the training set. Right: The self-driving car (Solid) and two nearby vehicles (Outlined). All motion is from left to right. The self-driving vehicle must forecast the trajectory of the nearby vehicels based on their observed path and any knowledge of the driver behaviour learned from the training set vehicles}
\label{fig:problem}
\end{figure}

@@

Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between the different sets of vehicles. These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent heterogenous (IH) model and the hierarchical, clustered heterogeneity (CH) model. The ID model does not incorporate information from the vehicles in the training into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set. 
\\

@@

The framework provided by (\ref{xEq}) and (\ref{yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \dots, T+H\}$, to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \dots, T+H\}$.  
\\

@@

These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and angle  $\delta_t$ are employed, given by
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
\end{align}
for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution. The unconditional means for acceleration and steering and are set to zero and $\pi/2$ respectively, corresponding to forward motion at a constant velocity.
\\

@@

The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them. To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous} (IH) model and a dependent \textit{clustered heterogeneous} (CH) model. In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})\}$ corresponding to the parameters given in the model for car $i$ given in
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{aAR2} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{dAR2}
\end{align}
with $\epsilon_{i,t}$ and $\eta_{i,t}$ all independent standard normal random variables. For the IH model, each $\theta_i,$ for $i=1,2,\ldots, N$, is assumed to follow an independent $N(\mu, \Sigma)$ prior distribution, resulting in a joint posterior distribution that may be decomposed into $N$ distinct independent marginal distributions, i.e.
\begin{equation}
p(\theta_{1:N} \mid z_{1:N}) = \prod_{i=1}^N p(\theta_{i} \mid z_{i}),
\end{equation}
due to the distinct parameters and independent (albeit identical) prior distributions for each of the $N$ models. \footnote{Recall that $z_{i}=\{x^*_{i,s},y^*_{i,s},\mbox{ for } s=1,2,...,T \}$.}
\\

In contrast, the alternative CH model does not assume \textit{a priori} independence of each of the $\theta_i$ vectors and instead clusters individual $\theta_i$ values into one of $K$ common latent multivariate normal distributions as determined by $\beta.$ The hierarchical CH model assumes that $\beta$ is itself random, with $\beta \sim p(\beta)$. By augmenting each driver's individual parameter $\theta_i$ with the corresponding $k_i=j$ variable, where $j$ can take on any value $1,2 \ldots, K$, we have
\begin{equation}
\label{mixPrior}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
\end{equation}
and
\begin{equation}
k_i \mid \beta \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
\end{equation}
for each $i = 1, \dots, N$. The mixture indicator variables satisfying $k_i=j$ determine a cluster of similar $\theta_i$ parameters corresponding to the behaviours of a subset of drivers. $\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, \mbox{ for } j=1,\},$ with \textit{a priori} assumptions for each cluster given by  
\begin{align}
\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right).
\end{align}
\\

@@

For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles in the training set, the vehicles where information is available for before any self-driving vehicle enters the road. Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional vehicle in the test set that may be encountered while the self-driving vehicle is on the road. Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available before driving.
\\

@@

Combining a fast inference procedure with the test vehicle posterior distributions implied by either of the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.

@@

Given a series of data for vehicle $i$ observed up to time $T$, $z_{i, 1:T}$, and the training set data $\textbf{z}_{1:N}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
\begin{equation}
\label{predictive}
p(z_{i, T+h} | z_{i, 1:T}) =\int p(z_{i, T+h}|z_{i, 1:T}, \theta_i) p(\theta_i | \textbf{z}_{1:N}, z_{i, 1:T}) d\theta.
\end{equation}
To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewCar2}) for the CH model must first be inferred. However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive}) is unavailable.
\\

Collectively denoting all available data, including both $\textbf{z}_{1:N}$ and $z_{i, 1:T}$, as $\textbf{z}$ for the remainder of this section, two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB). In brief, MCMC is used to create a sample from $p(\theta_i | \textbf{z})$, with any function of $\theta_i$ that is desired estimated from that sample. In contrast, VB replaces $p(\theta_i | \textbf{z})$ with a parametric approximation, denoted by $q_{\lambda}(\theta_i |\textbf{z})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $\textbf{z}$.

@@

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler. The Gibbs sampler algorithm iteratively samples the components of a $k$-dimensional parameter vector $\theta_i$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_{i, 1} | \theta_{i, 2}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
&p(\theta_{i, 2} | \theta_{i, 1}, \theta_{i, 3}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_{i, k} | \theta_{i, 1}, \dots, \theta_{i, k-1}, \textbf{z}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta_i | \textbf{z})$. Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm. The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically increases with the amount of data and number of parameters in the model. The full conditional distributions cannot be recognised for each of the models proposed in Section \ref{sec:models}, and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
\\

@@

Eucliedean Error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}. Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error. Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3). At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model. The average forecast Euclidean Error for the best performing Naive model, Naive 8, and each of the time series models at each of these three forecast horizons are presented in Table \ref{table:meanError}.
\\
